Translation of B Implementations to the LLVM: Control and Modularity

### 2013-08-22 translation.tex
Created file translation.tex to include all translation rules, and the description of the validation. Added an example used for manual translation. Created a prototype implementation of the translation in Python. Added a pythonesque representation of the example suitable for processing by the prototype.

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Control and Modularity}
\author{David Déharbe \and Valério Medeiros Jr} 
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\name}{\mathcal{L}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

        \B{init} & : & \B{Inst} \\

@@

        \B{lhs} & : & \B{Name} \\

@@

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{struct} \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity.}
  \label{fig:llvm-grammar}
\end{figure}

@@

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to LLVM constants. To avoid
  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
\item To encode the implementation state, the translation produces a
  structure type, named \llvm{@M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines
  are encoded as elements of that structure.  Also, for each operation of each
  imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

  The parameter list of the functions thus produced contain one item for each
  input and output of the corresponding operation, and one item of type
  \llvm{@M\$state\$*}, to pass the address of the structure encoding the
  state of the implementation and its modification by the callee. The return
  type is always \llvm{void}.
\end{enumerate}
This approach assumes that information on external B components used in \B{M} is
available during translation. To fulfill this hypothesis, the translation
produces an ``interface'' artifact containing the information that the
translation of an importation of \B{M} in a client machine
requires. Figure~\ref{fig:skel} summarizes the translation principles by
providing the structure of both the interface and the LLVM implementation thus
produced.

\begin{figure}
  \begin{center}
    (a) Structure of the interface artifact produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{(references to imported components)} \\
      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
      \llvm{declare void @M\$init\$(@M\$state\$*)} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{declare void @M\$op(@M\$state\$*, \ListOf{\nt{type}})} \\
    \end{tabbing}
    (b) Structure of the LLVM module produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{@Q\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(@Q\$state\$*)} \\
      \> \textit{for each constant} \B{cst} \textit{generate} \\
      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(@Q\$state\$*, \ListOf{\nt{type}})} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
      \llvm{@M\$self\$ = common global @M\$state\$ zeroinitializer} \\
      \llvm{define void @M\$init\$(@M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(@M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}}
    \end{tabbing}
  \end{center}
  \caption{Skeleton of the artifacts produced by the translation.}
  \label{fig:skel}
\end{figure}

@@

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name. Also $\state(\B{M})
\isdef \llvm{"@\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
state of component \B{M}. We assume an unlimited supply of labels and local
variables: let $\name$ return a fresh LLVM label or local variable name.

@@

\item $\intf(\B{M})_t$ is a textual description of the type representing the
  state of the component, i.e. a structure with one element per imported
  component and variable:
\begin{small}
$$\intf(\B{M})_t \isdef \llvm{"struct \{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
\end{small}
where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.

@@

B constants are translated to constants in the LLVM program, as described in the
following rule, where parameter \B{c} denote the constant, $M$ the name of the
module and $\Global$ is the mapping between source and target identifiers. Two
results are produced: the LLVM source code and an updated identifier mapping:
\begin{small}
\begin{align*}
  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
\end{align*}
\end{small}
The definition of the translation for expressions is described next in
section~\ref{sec:expr}.

@@

@@

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

\paragraph{Comparisons} are formed by the application of a relational operator
to two expressions. Mapping $\opmap$ associates B operators to corresponding
LLVM keywords:
$$\opmap = \{ = \mapsto \llvm{"eq"}, 
  \ne \mapsto \llvm{"ne"}, 
  < \mapsto \llvm{"slt"}, 
  \le \mapsto \llvm{"sle"}, 
  > \mapsto \llvm{"sgt"}, 
  \ge \mapsto \llvm{"sge"} \}.
$$

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \AND v = \name \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations} have a similar structure and interpretation
as comparisons and their presentation is mostly omitted. Let us present the
translation of the unary \B{succ} operator though:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

@@

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

@@

Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
of B instructions \B{il} and a label $\ell$ and produces the source code of a
LLVM block with $\ell$ as exit point. It is defined as follows:

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Control and Modularity}
\author{David Déharbe \and Valério Medeiros Jr} 
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

\newcommand{\DD}[1]{\marginpar{\tiny{DD: #1}}}

@@

        \B{init} & : & seq \B{Inst} \\

@@

        \B{lhs} & : & \B{Vari} \\

@@

\DD{Include \B{Vari} as one of the alternative for an \B{Expr}.}
\DD{Restore \B{Case} and \B{CaseBr}.}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity.}
  \label{fig:llvm-grammar}
\end{figure}

@@

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to LLVM constants. To avoid
  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
\item To encode the implementation state, the translation produces a
  structure type, named \llvm{\%M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines
  are encoded as elements of that structure.  Also, for each operation of each
  imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

  The parameter list of the functions thus produced contain one item for each
  input and output of the corresponding operation, and one item of type
  \llvm{\%M\$state\$*}, to pass the address of the structure encoding the
  state of the implementation and its modification by the callee. The return
  type is always \llvm{void}.
\end{enumerate}
This approach assumes that information on external B components used in \B{M} is
available during translation. To fulfill this hypothesis, the translation
produces an ``interface'' artifact containing the information that the
translation of an importation of \B{M} in a client machine
requires. Figure~\ref{fig:skel} summarizes the translation principles by
providing the structure of both the interface and the LLVM implementation thus
produced.

\begin{figure}
  \begin{center}
    (a) Structure of the interface artifact produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{(references to imported components)} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{declare void @M\$op(\%M\$state\$*, \ListOf{\nt{type}})} \\
    \end{tabbing}
    (b) Structure of the LLVM module produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
      \> \textit{for each constant} \B{cst} \textit{generate} \\
      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer} \\
      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}}
    \end{tabbing}
  \end{center}
  \caption{Skeleton of the artifacts produced by the translation.}
  \label{fig:skel}
\end{figure}

@@

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name. Also $\state(\B{M})
\isdef \llvm{"\%\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
state of component \B{M}. We assume an unlimited supply of labels and local
variables: we use $\lbl$ to return a fresh LLVM label and $\variable$ for a
fresh local variable name.

@@

\item $\intf(\B{M})_t$ is a textual description of the type representing the
  state of the component, i.e. a structure with one element per imported
  component and variable:
\begin{small}
$$\intf(\B{M})_t \isdef \llvm{"\{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
\end{small}
where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.

@@

B constants are translated to constants in the LLVM program, as described in the
following rule, where parameter \B{c} denote the constant, $M$ the name of the
module and $\Global$ is the mapping between source and target identifiers. Two
results are produced: the LLVM source code and an updated identifier mapping:
\begin{small}
\begin{align*}
  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
\end{align*}
\end{small}

The definition of the translation for expressions is described next in
section~\ref{sec:expr}.

@@

To facilitate the defintiion of rules that apply to similar cases, we introduce
mapping $\opmap$, that associates B operators to corresponding LLVM keywords:
$$\opmap = \{ 
\begin{array}[t]{l}
  = \mapsto \llvm{"eq"}, 
  \ne \mapsto \llvm{"ne"}, 
  < \mapsto \llvm{"slt"}, 
  \le \mapsto \llvm{"sle"}, 
  > \mapsto \llvm{"sgt"}, 
  \ge \mapsto \llvm{"sge"}, \\
  + \mapsto \llvm{"add"} \}.
  \end{array}
$$

@@

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

\paragraph{Comparisons} are formed by the application of a relational operator
to two expressions.

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \IN \LET v = \variable \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations}\DD{Actually some arithmetic operations are not binary and different translation rules shall be written.} have a similar 
structure and interpretation as comparisons. Here is the rule for binary
arithmetic operations:
\begin{align*}
\lefteqn{\trad{r}{\B{Term}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \IN \LET v = \variable \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = i32 \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v \sep \llvm{i32}
\end{align*}

Next is the translation of the unary \B{succ} operator:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
  \trad{n.arg}{\B{expr}} \IN \LET w = \variable \IN \\
  & \llvm{"\PH{p} \quad \PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

@@

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \lbl \IN} \\
  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

@@

Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
of B instructions \B{il} and a label $\ell$ and produces the source code of a
LLVM block with $\ell$ as exit point. It is defined as follows: \DD{Rewrite so
  that the temporary variable names are generated in the correct order.}
